Embedded Program Search in Agda

### 2020-05-06 paper/paper.tex
final commit

Modules need to be redefined(See Figure \ref{fig:mods}) from the abstraction in
the previous section because they allow for codes to reference other
definitions. We will define Module similarly to the type synonym we used above,
except that it will be mutually defined with the interpretation function. Note
that for the $\_::\_$ case we provide a module first in the constructor so that
the entry may depend on it. The entries in the module will be similar to that in
the example module in Section 4. In the same way the dependent pair
did in the previous Module definition.

@@

To make all of this a bit more concrete, Figure \ref{fig:ex} is an example of a
module encoded into this typing universe. Here each entry is preceded by its
type in the coded type system. Note that the references properly index the types
of their functions or terms.

@@

This formalism thus far is lacking ergonomics. The tool is rather impractical
because the module encoding needs to be written by hand in a coded language that
relies heavily on numerical indexing. We alleviate this problem with a set of
programs which both generate the coded modules and the coded queries in
\textsf{TypeCode}. These two programs hook into the Agda Compiler to parse the
target library files and generate an auxiliary file containing the coded modules
and queries. The queries will need to be provided as codes in \textsf{TypeCode}.
The user provides the syntax valid for Agda and the tool generates the element
of \textsf{TypeCode} corresponding to the type provided by converting its base
types to the proper references and function arrows to the coded-arrows.

Along with these programs we have created an integration with emacs to run the
indexing program whenever emacs saves an Agda file. The query executable can be
called using a shortcut or from the list of available emacs commands. We have
found that the inclusion of these features made the search procedure practical
for the end-user.

@@

\paragraph{Typing Universes} There has been much work on universes for generic
programming. Benke et al. create a collection of category-inspired universes
specifically with the aim of formalizing past work on generic programming in the
vein of Generic Haskell \cite{benke2003universes}. This work differs from ours
in its aim, it mainly attempts to automate the writing of functions such as
generic map or fold which have a common behavior no matter the structure of the
type. We attempt to be generic over the types themselves but with a more
concrete aim. This makes this approach to abstract for our use. Another line of
work aims at formalizing type theory in type theory using quotient inductive
types\cite{altenkirch2016type}. This work is closer to our own in that they
encode their universe mutually in a similar way to our encoding. However this
work aims at verifying type theory within itself and makes heavy use of Homotopy
Type Theory\cite{HoTT}. The formalism itself is not quite adequate for our
purposes as it lacks an mechanism for interpreting its type encodings.

@@

Supporting the generation of first class modules from multiple files is also a
near term improvement that is technically not too difficult. The name spaces
will need to be qualified and import conflicts resolved. This will open up the
practical use of our tool for searching over a large repository of code such as
the Agda Standard Library.\cite{bove2009brief}

@@

One of the items lacking in this project was a successful encoding of a universe
for fully dependent types. There was a prototype produced that had some
deficiencies preventing it from having a sound interpretation function. This
will be the first follow-up work to be conducted. Completing this universe will
let the search procedure handle example-based searches such as was show in the
overview section. This need not stop at examples however. Any decidable property
over terms/functions can be used in the search process. Thus this formalism will
be much stronger than any program search tools currently available for
functional languages.

@@

A key realization motivating this line of research is the relation of program
search to the more general problem of program synthesis. Program Synthesis is
the task of generating a program which satisfies some specification. Types
provide one such specification. If the synthesis task is then reduced to
producing some term of a given query type, the synthesis procedure must produce
an abstract syntax tree that typechecks to the query type. If we restrict the
solution to have an AST of size 1 it essentially forces the synthesis problem to
become program search.

This is not a particularly profound idea. However, when considering program
search as foundational operation in synthesis procedures it becomes more
motivating. This was the reasoning behind formalizing search within Agda as
opposed to as a separate tool. With a function we can compose search with other
functions to build a complex framework for synthesis over dependent types. This
is our main line of future work in this area.

>>>>

Modules need to be redefined(See Figure \ref{fig:mods}) from the abstraction in
the previous section because they allow for codes to reference other
definitions. We will define Module similarly to the type synonym we used above,
except that it will be mutually defined with the interpretation function. Note
that for the $\_::\_$ case we provide a module first in the constructor so that
the entry may depend on it. The entries in the module will be similar to that in
the example module in Section 4. This is in the same way the dependent pair
did in the previous Module definition.

@@

More concretely, Figure \ref{fig:ex} is an example of a module encoded into this
typing universe. Here each entry is preceded by its type in the coded type
system. Note that the references properly index the types of their functions or
terms.

@@

This formalism thus far is lacking in ergonomics. The tool is rather impractical
because the module encoding needs to be written by hand in a coded language that
relies heavily on numerical indexing. We alleviate this problem with a set of
programs which both generate the coded modules and the coded queries in
\textsf{TypeCode}. These two programs hook into the Agda Compiler to parse the
target library files and generate an auxiliary file containing the coded modules
and queries. The queries will need to be provided as codes in \textsf{TypeCode}.
The user provides the syntax valid for Agda and the query tool generates the
element of \textsf{TypeCode} corresponding to the type provided by converting
its base types to the proper references and function arrows to the coded-arrows.

Along with these programs we have created an integration with Emacs to run the
indexing program whenever Emacs saves an Agda file. The query executable can be
called using a shortcut or from the list of available emacs commands. We have
found that the inclusion of these features makes the search procedure practical
for the end-user.

@@

\paragraph{Typing Universes} There has been much work on universes for generic
programming. Benke et al. create a collection of category-inspired universes
specifically with the aim of formalizing past work on generic programming in the
vein of Generic Haskell \cite{benke2003universes}. This work differs from ours
in its aim, it mainly attempts to automate the writing of functions such as
generic map or fold which have a common behavior no matter the structure of the
type. We attempt to be generic over the types themselves but with a more
concrete aim. This approach is too abstract for our use. Another line of work
aims at formalizing type theory in type theory using quotient inductive
types\cite{altenkirch2016type}. This work is closer to our own in that it
encodes the universe mutually. However this work aims at verifying type theory
within itself and makes heavy use of Homotopy Type Theory\cite{HoTT}. The
formalism itself is not quite adequate for our purposes as it lacks a mechanism
for interpreting its type encodings.

@@

Supporting the generation of first class modules from multiple files is also a
near-term improvement that is technically not too difficult. The namespaces
will need to be qualified and import conflicts resolved. This will open up the
practical use of our tool for searching over a large repository of code such as
the Agda Standard Library\cite{bove2009brief}.

@@

One of the items lacking in this project was a successful encoding of a universe
for fully dependent types. There was a prototype produced that had some
deficiencies preventing it from having a sound interpretation function. This
will be the first follow-up work to be conducted. Completing this universe will
let the search procedure handle example-based searches such as was shown in the
overview section. This need not stop at examples. Any decidable property over
terms or functions can be used in the search process. Thus this formalism will
be much stronger than any program search tools currently available for
functional languages.

@@

A key realization motivating this line of research is the relation of program
search to the more general problem of Program Synthesis. Program Synthesis is
the task of generating a program which satisfies some specification. Types
provide one such specification. If the synthesis task is then reduced to
producing some term of a given query type, the synthesis procedure must produce
an abstract syntax tree that typechecks to the query type. If we restrict the
solution to have an AST of size 1 it essentially forces the synthesis problem to
become program search.

This is not a particularly profound idea. However, when considering program
search as a foundational operation in synthesis procedures it becomes more
interesting. This was the reasoning behind formalizing search within Agda as
opposed to in a separate tool. With a function we can compose search with other
functions to build a complex framework for synthesis over dependent types. This
is our main line of future work.
